---
title: "eQTL mapping with tQTLExperiment and tensorQTL"
author: "Vincent Carey"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{eQTL mapping with tQTLExperiment and tensorQTL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`tQTLExperiment` wraps the
[tensorQTL](https://github.com/broadinstitute/tensorqtl) GPU-accelerated eQTL
mapping tool in a
[SummarizedExperiment](https://bioconductor.org/packages/SummarizedExperiment)-style
container.  The key design principle is a **two-step workflow** that keeps R and
Python in separate environments:

1. `prepareTQTL()` — writes phenotype and covariate files to a directory and
   returns the complete `python -m tensorqtl` command string.
2. The user runs that command in a terminal with tensorQTL installed.
3. `readTQTL()` — reads the output files back into R as
   [GRanges](https://bioconductor.org/packages/GenomicRanges).

This avoids all R/Python environment conflicts (conda PATH issues, `rpy2`
segfaults, duplicate OpenMP runtimes).

Genotype data are represented **lazily** via
[BEDMatrix](https://CRAN.R-project.org/package=BEDMatrix) — the PLINK `.bed`
file is never loaded into R memory.  This makes it practical to work with
whole-genome genotype files containing millions of variants.

## Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("vjcitn/tQTLExperiment")
```

tensorQTL must be installed in a Python environment:

```bash
pip install tensorqtl
```

# MAGE example data

This vignette uses data from the
[MAGE](https://www.science.org/doi/10.1126/science.adi6529) study
(Multi-Ancestry Genomics, Epigenomics, and Transcriptomics), a multi-ancestry
eQTL resource.  The `CSHLvc2026` companion package provides cached access to
chromosome 17 genotypes and a pre-filtered pseudobulk B-cell SummarizedExperiment.

```{r loaddata, eval=FALSE}
library(tQTLExperiment)
library(CSHLvc2026)

data(mageSEfilt, package = "CSHLvc2026")
mageSEfilt
```

```{r echo=FALSE}
library(tQTLExperiment)
```

The genotype data live in PLINK format on OSN cloud storage and are downloaded
once via `BiocFileCache`:

```{r getgeno, eval=FALSE}
plink_paths <- CSHLvc2026::cache_mage_chr17_plink()
plpre <- tools::file_path_sans_ext(plink_paths[["bed"]])
```

# Building a tQTLExperiment

## Covariate matrix

tensorQTL requires numeric covariates.  We build a model matrix from `colData`
using `model.matrix()`, then drop the intercept column — tensorQTL does not
expect one.  Dropping the first column after construction (rather than using
`~ 0 + ...`) preserves reference-level dummy coding for factor variables.

```{r covmat, eval=FALSE}
cd <- as.data.frame(colData(mageSEfilt))
mm <- model.matrix(~ batch + population + sex, data = cd)
mm <- mm[, -1, drop = FALSE]   # remove (Intercept)
```

## Constructor

`tQTLExperimentFromRSE()` takes an existing
`RangedSummarizedExperiment` and PLINK prefix, attaches the lazy genotype
matrix, and optionally records the genome build in `seqinfo`.

```{r construct, eval=FALSE}
tqe <- tQTLExperimentFromRSE(
    se              = mageSEfilt,
    plinkPrefix     = plpre,
    covariateMatrix = mm,
    genome          = "hg38"
)
tqe
```

## Adding gene symbols

`addGeneSymbols()` looks up HGNC names via `EnsDb.Hsapiens.v79` and adds a
`gene_name` column to `mcols(rowRanges(tqe))`.  Ensembl version suffixes
(e.g. `.3`) are stripped automatically.

```{r syms, eval=FALSE}
tqe <- addGeneSymbols(tqe)
head(mcols(rowRanges(tqe))[, c("phenotype_id", "gene_name")])
```

# Running tensorQTL

## Subset to chr17 features

The PLINK file covers chromosome 17 only, so we filter the SE to matching
features before writing the phenotype file.

```{r subset, eval=FALSE}
is_17 <- as.character(seqnames(rowRanges(tqe))) == "chr17"
tqe17 <- tqe[which(is_17), ]
dim(tqe17)
```

## prepareTQTL

`prepareTQTL()` writes `pheno.bed` and `covariates.tsv` to `outDir` and
returns the shell command.  The built-in `--maf_threshold` flag is one of
tensorQTL's key advantages — no pre-filtering step is required.

```{r prepare, eval=FALSE}
outdir <- "~/mage_tqtl_run"
dir.create(outdir, showWarnings = FALSE)

cmd <- prepareTQTL(
    tqe17,
    outDir       = outdir,
    mode         = "cis_nominal",
    mafThreshold = 0.05
)
cat(cmd, "\n")
```

The printed command looks like:

```
python3 -m tensorqtl /path/to/CCDG_mage_chr17 ~/mage_tqtl_run/pheno.bed \
    ~/mage_tqtl_run/tqtl_out --mode cis_nominal --maf_threshold 0.05 \
    --window 1000000 -o ~/mage_tqtl_run \
    --covariates ~/mage_tqtl_run/covariates.tsv
```

Paste this into a terminal where your tensorQTL conda environment is active:

```bash
conda activate plink-env
KMP_DUPLICATE_LIB_OK=TRUE <paste command>
```

## readTQTL

Once tensorQTL finishes, load results back into R.  Passing `x = tqe17`
propagates the genome build and gene symbols automatically.

```{r read, eval=FALSE}
res <- readTQTL(outdir, mode = "cis_nominal", x = tqe17)
res$pairs
```

# Exploring results

```{r explore, eval=FALSE}
pairs <- res$pairs

# top associations by nominal p-value
pairs[order(pairs$pval_nominal)][1:10,
      c("gene_name", "variant_id", "af", "slope", "pval_nominal")]

# associations for a specific gene
tp53bp1 <- pairs[pairs$gene_name == "TP53BP1" & !is.na(pairs$gene_name)]
tp53bp1[order(tp53bp1$pval_nominal)][1:5]

# distribution of nominal p-values
hist(pairs$pval_nominal, breaks = 50,
     main = "tensorQTL nominal p-values (chr17, cis)",
     xlab = "p-value")
```

# Example data (chr22)

The package ships with a small example dataset (20 genes, 100 samples,
chromosome 22) that can be used without external data or a tensorQTL
installation to verify the object construction steps.

```{r example}
exdir <- system.file("extdata", package = "tQTLExperiment")

tqe_ex <- tQTLExperiment(
    plinkPrefix = file.path(exdir, "chr22-n100"),
    phenoFile   = file.path(exdir, "mean-pheno-n100.bed"),
    genome      = "hg38"
)
tqe_ex
rowRanges(tqe_ex)
tqtlGeno(tqe_ex)
```

The tensorQTL command for this example would be obtained with:

```{r prepcmd}
od  <- tempdir()
cmd <- prepareTQTL(tqe_ex, outDir = od, mode = "cis_nominal")
cat(cmd, "\n")
```

# Session info

```{r sessioninfo}
sessionInfo()
```
